### 2018-02-15 submission/manuscript.Rmd
update discussion section

Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with tumors versus controls [Figure 1-3]. Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect size increases in OR for either adenoma or carcinoma tumors. Even though mouth-associated genera increased individaul's OR of having a carcinoma, they did not consistently increase the OR of having an adenoma. Additionally, our observations suggest that by combining the top 5 most protective and detrimental microbes we can classify indviduals with either adenoma or carcinoma as well as models that use the full community.

The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, tumor formation. The results that we have presented show that both the genera- and OTU-based select and full models, for indviduals with carcinoma, had similar AUCs [Figure 2 & 3]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community could be crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that should be studied further to potentially help reduce the risk of carcinoma tumor formation. Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma turmors. Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. Additionally, how the host interacts with these subtle changes at early stages of the disease could be what leads to a thoroughly dysfunctional community that is supportive of tumorgenesis.

Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 4]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both individuals with adenoma and carcinoma [Figure 4]. These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.

@@

One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. Since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies. This suggests that even though *Bacteroides* may not increase the OR of individuals having an adenoma or carcinoma and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC. Additionally, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased OR for individuals to have either an adenoma or carcinoma.

The associations between the microbiota and indivdials with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and those with adenomas. Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to polyp formation. In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 5].

@@

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in colorectal tumors. The data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are commonly associated with the greatest OR of individuals having a carcinoma. Conversely, no conclusive signal with these mouth-associated microbes could be detected for individuals with an adenoma. Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma tumors due to possible subtle changes in the community. Overall, associations between the microbiota and individuals with carcinomas were much stronger than with those with adenomas.

>>>>

Our study identifies clear but small differences in diversity at the community level and larger differences for individual genera, present in patients with tumors versus controls [Figure 1-3]. Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect size increases in OR for either adenoma or carcinoma tumors. Even though mouth-associated genera increased individaul's OR of having a carcinoma for certain sample types, they did not consistently increase the OR of having an adenoma. By using these taxa that had significant ORs after multiple comparison correction we found that we could classify indviduals with either adenoma or carcinoma as well as models that use either all genera or all OTUs. Finally, many studies were individually under powered to be able to reject the null hypothesis and this could one reason only the comparison between control and carcinoma individuals for stool samples had relible detectable differences.

The data presented herein support the importance of specific taxa for carcinoma, but not necessarily adenoma, tumor formation. The results that we have presented show that the significant OR taxa model and both the full genera and OTU models, for indviduals with carcinoma, had similar AUCs [Figure 2 & 3]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community could be crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that should be studied further to potentially help reduce the risk of carcinoma tumor formation. Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma tumors. Regardless of sample type and whether a full genera- or OTU-based model was used, our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. How the host interacts with these subtle changes at early stages of the disease could be what leads to a thoroughly dysfunctional community that is supportive of tumorgenesis.

For the full genera- and OTU-based models within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 4]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both individuals with adenoma and carcinoma [Figure 4]. These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated or general inflammatory genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.

@@

One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. Since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies. This suggests that even though *Bacteroides* may not increase the OR of individuals having an adenoma or carcinoma and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC. Additionally, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased OR for individuals to have a carcinoma tumor.

The associations between the microbiota and individuals with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and those with adenomas. Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to adenoma formation. In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 5].

@@

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in colorectal tumors. The data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are commonly associated with the greatest OR of individuals having a carcinoma. Conversely, no conclusive signal with these mouth-associated microbes could be detected for individuals with an adenoma. Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma tumors due to possible subtle changes in the community. Overall, the associations between the microbiota and individuals with carcinomas were much stronger than with those with adenomas.

### 2018-02-15 submission/manuscript.tex
update discussion section

Our study identifies clear differences in diversity, both at the
community level and for individual genera, present in patients with
tumors versus controls {[}Figure 1-3{]}. Although there was a step-wise
decrease in diversity as disease progressed from control to adenoma to
carcinoma, this did not translate into large effect size increases in OR
for either adenoma or carcinoma tumors. Even though mouth-associated
genera increased individaul's OR of having a carcinoma, they did not
consistently increase the OR of having an adenoma. Additionally, our
observations suggest that by combining the top 5 most protective and
detrimental microbes we can classify indviduals with either adenoma or
carcinoma as well as models that use the full community.

The data presented herein support the importance of select genera for
carcinoma, but not necessarily adenoma, tumor formation. The results
that we have presented show that both the genera- and OTU-based select
and full models, for indviduals with carcinoma, had similar AUCs
{[}Figure 2 \& 3{]}. This suggests that an interplay between a select
number of potentially protective and exacerbating microbes within the GI
community could be crucial for carcinoma formation. Importantly, it
suggests that there may be key members of the GI community that should
be studied further to potentially help reduce the risk of carcinoma
tumor formation. Conversely, using the present data, it is clear that
new approaches may be needed to identify members of the community
associated with adenoma turmors. Regardless of sample type and whether a
full or select model was used, our Random Forest models consistently
performed poorly. Yet, the step-wise decrease in diversity suggests that
the adenoma-associated community is not normal but has changed subtly.
This change in diversity, at this early stage of disease, could be focal
to the adenoma itself. Additionally, how the host interacts with these
subtle changes at early stages of the disease could be what leads to a
thoroughly dysfunctional community that is supportive of tumorgenesis.

Within stool, common GI microbes were most consistently present in the
top 10 genera or OTUs across studies {[}Figure 4{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be in the top 10 most
important variables across the different studies for both individuals
with adenoma and carcinoma {[}Figure 4{]}. These data suggest that
whether the non-resident bacterium is \emph{Fusobacteria} or
\emph{Peptostreptococcus} may not be as important as how these bacteria
interact with the changing resident community. Based on these
observations, it is possible to hypothesize that small changes in
community structure lead to new niches in which any one of the
mouth-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma stage of disease.

@@

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
OR of individuals having an adenoma or carcinoma and may not vary in
relative abundance, like \emph{Fusobacterium}, it is still important in
CRC. Additionally, \emph{Streptococcus gallolyticus} subsp.
\emph{gallolyticus} is a mouth-associated microbe, and the results from
this study suggest that regardless of sample type, mouth-associated
genera are commonly associated with an increased OR for individuals to
have either an adenoma or carcinoma.

The associations between the microbiota and indivdials with adenoma
tumors are inconclusive, in part, because many studies may not be
powered effectively to observe small effect sizes. None of the studies
analyzed were properly powered to detect a 10\% or lower change between
cases and controls. The results within our meta-analysis suggest that a
small effect size may well be the scope in which differences
consistently occur between controls and those with adenomas. Future
studies investigating adenoma tumors and the microbiota need to take
power into consideration to reproducibly study whether the microbiota
contributes to polyp formation. In contrast to adenoma stage of disease,
our observations suggest that most studies analyzed have sufficient
power to detect many changes in the carcinoma-associated microbiota
because of large effect size differences between cases and controls
{[}Figure 5{]}.

>>>>

Our study identifies clear but small differences in diversity at the
community level and larger differences for individual genera, present in
patients with tumors versus controls {[}Figure 1-3{]}. Although there
was a step-wise decrease in diversity as disease progressed from control
to adenoma to carcinoma, this did not translate into large effect size
increases in OR for either adenoma or carcinoma tumors. Even though
mouth-associated genera increased individaul's OR of having a carcinoma
for certain sample types, they did not consistently increase the OR of
having an adenoma. By using these taxa that had significant ORs after
multiple comparison correction we found that we could classify
indviduals with either adenoma or carcinoma as well as models that use
either all genera or all OTUs. Finally, many studies were individually
under powered to be able to reject the null hypothesis and this could
one reason only the comparison between control and carcinoma individuals
for stool samples had relible detectable differences.

The data presented herein support the importance of specific taxa for
carcinoma, but not necessarily adenoma, tumor formation. The results
that we have presented show that the significant OR taxa model and both
the full genera and OTU models, for indviduals with carcinoma, had
similar AUCs {[}Figure 2 \& 3{]}. This suggests that an interplay
between a select number of potentially protective and exacerbating
microbes within the GI community could be crucial for carcinoma
formation. Importantly, it suggests that there may be key members of the
GI community that should be studied further to potentially help reduce
the risk of carcinoma tumor formation. Conversely, using the present
data, it is clear that new approaches may be needed to identify members
of the community associated with adenoma tumors. Regardless of sample
type and whether a full genera- or OTU-based model was used, our Random
Forest models consistently performed poorly. Yet, the step-wise decrease
in diversity suggests that the adenoma-associated community is not
normal but has changed subtly. This change in diversity, at this early
stage of disease, could be focal to the adenoma itself. How the host
interacts with these subtle changes at early stages of the disease could
be what leads to a thoroughly dysfunctional community that is supportive
of tumorgenesis.

For the full genera- and OTU-based models within stool, common GI
microbes were most consistently present in the top 10 genera or OTUs
across studies {[}Figure 4{]}. Changes in \emph{Bacteroides},
\emph{Ruminococcaceae}, \emph{Ruminococcus}, and \emph{Roseburia} were
consistently found to be in the top 10 most important variables across
the different studies for both individuals with adenoma and carcinoma
{[}Figure 4{]}. These data suggest that whether the non-resident
bacterium is \emph{Fusobacteria} or \emph{Peptostreptococcus} may not be
as important as how these bacteria interact with the changing resident
community. Based on these observations, it is possible to hypothesize
that small changes in community structure lead to new niches in which
any one of the mouth-associated or general inflammatory genera can gain
a foothold, exacerbating the initial changes in community and
facilitating the transition from adenoma to carcinoma stage of disease.

@@

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
OR of individuals having an adenoma or carcinoma and may not vary in
relative abundance, like \emph{Fusobacterium}, it is still important in
CRC. Additionally, \emph{Streptococcus gallolyticus} subsp.
\emph{gallolyticus} is a mouth-associated microbe, and the results from
this study suggest that regardless of sample type, mouth-associated
genera are commonly associated with an increased OR for individuals to
have a carcinoma tumor.

The associations between the microbiota and individuals with adenoma
tumors are inconclusive, in part, because many studies may not be
powered effectively to observe small effect sizes. None of the studies
analyzed were properly powered to detect a 10\% or lower change between
cases and controls. The results within our meta-analysis suggest that a
small effect size may well be the scope in which differences
consistently occur between controls and those with adenomas. Future
studies investigating adenoma tumors and the microbiota need to take
power into consideration to reproducibly study whether the microbiota
contributes to adenoma formation. In contrast to adenoma stage of
disease, our observations suggest that most studies analyzed have
sufficient power to detect many changes in the carcinoma-associated
microbiota because of large effect size differences between cases and
controls {[}Figure 5{]}.
